Time stretching of the GeV emission of GRBs: Fermi LAT data vs geometrical model

### 2014-12-12 Paper/gammaRays.tex
Fixed a few minor typos

We start with the location, time and duration of bursts from the
Fermi-LAT Gamma-Ray Burst Catalog \cite{Ackermann:2013zfa}. The
photons during the burst and the photons for 24 hours before the burst
are downloaded from the LAT Data
Server~\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}.
The latter are used for the background estimation in both energy bands
with the technique introduced in \cite{Rubtsov:2011qq}. In order to
have enough statistics we require that at least 10 photons are
detected with the energy greater than $\unit[1]{GeV}$. This leaves are
with the four bursts for the time span of the Fermi-LAT Gamma-Ray
Burst Catalog. Namely, GRB~080916C \cite{Tajima:2009az}, GRB~090510
\cite{Ackermann:2010us}, GRB~090902B \cite{Abdo:2009pg} and
GRB~090926A \cite{Bregeon:2011bu}. For these bursts we stretch the
high energy light curve by the arbitrary factor and compare with the
low energy light curve using the Kolmogorov-Smirnov (KS) test. Within
the $95\%$ confidence level the stretching factor is compatible to $1$
(no stretching) for the GRB 080916C and GRB 090510. For the other two
bursts the stretching factor of 1 is, however, excluded.  For GRB
090902B it should be smaller than 1, so the low energy light curve is
stretched with respect to the high energy one, see
fig. \ref{fig:grb090902B}.  Even more significant deviation from the
stretching factor of 1 in the opposite direction is observed for the
GRB 090926A. The high energy light curve of GRB 090926A is stretched
with respect to the low energy one by a factor of at least 1.99 (see
fig. \ref{fig:grb090926A}).  The stretching of the GRB 090926A high
energy light is found with $3.3\sigma$ significance.  All the observed
stretching factors are summarized in Table
\ref{tab:observationResults}, and the detailed procedure for their
calculation is described in Section \ref{sec:observations}.

@@

First, we download the Pass 7REP (V15) ``SOURCE'' class photons from
the Fermi-LAT data
server~\cite{2009ApJ...697.1071A,Ackermann:2012kna}. In order to
define the regions of interest we use the catalog
\cite{Ackermann:2013zfa}, which contains all the bright gamma-ray
bursts seen by the Fermi LAT since 2008 August 4 to 2011 August 1.
Specifically, we take 4 pieces of information from the tables 2 and 4
of Ref. \cite{Ackermann:2013zfa}:
\begin{itemize}
	\item{
		GBM trigger time.
		It is used as a reference point for the GRB time.
	}
	\item{
		Location of the GRB.
		The burst locations are used to filter out photons
                coming from the other sources.
	}
	\item{
		Location error.
		Location errors are used to improve the accuracy of filtering, specifically to avoid losing statistics by filtering too much.
	}
	\item{
		$T_{05}$ -- $T_{95}$ interval of the LAT-detected
          emission. The beginning of the interval is used as a fixed
          point of stretching. The Fermi LAT photons are downloaded
          for the period extended in time by $50\%$ towards both past
          and future compared to this interval.
	}
\end{itemize}
The data are summarized in table \ref{tab:bursts}. For the background
estimation, we also download observational data for 24 hours before
the burst.  The analysis is performed with the Fermi Science Tools
v9r32p5
package~\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/scitools/overview.html}}
following the
guidelines~\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/scitools/data_preparation.html}}.
We require Earth zenith angle to be below $100^\circ$ to remove the
Earth limb emission. After the basic filtering by \texttt{gtselect}
and \texttt{gtmktime} tools is done we perform more elaborate
selection using the point spread function of Fermi LAT, calculated
with \texttt{gtpsf} tool. We use 300 energy bins from \unit[100]{MeV}
to \unit[300]{GeV}, and 300 angular bins with maximal angle being
$30\degree$.  The \texttt{gtltcube} tool is used to compute required
livetime maps. Note, that we calculate PSFs separately for both
conversion types: front and back. We define $\mathrm{PSF}_{95}$ as an
angular distance from the source containing 95\% of the emitted
photons. We require that the photon is closer than $\mathrm{PSF}_{95}
+ \texttt{location.err}$ to the location of the GRB (see Table
\ref{tab:bursts}). The details of the data request and
analysis are given in appendix \ref{sec:fermiCode}.

@@

Although the details of the exact mechanism of the GRB are not known
the are arguments in support of our model.

@@

\subsection{Light curve}
We now know enough to compute the observable quantity - the number of
observed photons $p$. For that we need to integrate the radiation
intensity $\eta$ over the four groups of variables. Two groups are
burst-related: radiator position $\left(r, \theta, \phi\right)$ and
frequency at emission $\omega'$. The other two variables are related
to observer: observation time $\tau$ and frequency at observation
$\omega$. We relate the first two and the second two with the delta
functions:
\begin{align}
p\left( z, \chi; \tau_1, \tau_2; \omega_1, \omega_2 \right) &= \frac{A_\text{det}}{A_\text{ph}\left(z\right)} \int_0^\infty \dif r \int_0^{\pi} r \dif \theta \int_0^{2\pi} r \sin{\theta} \dif \phi \int_0^\infty \dif \omega' \int_{\tau_1}^{\tau_2} \dif \tau \int_{\omega_1}^{\omega_2} \dif \omega \, \eta\left( r, \theta, \omega' \right) \nonumber\\
&\qquad {} \times \underbrace{\frac{1}{\gamma^2\left( 1 - v \cos\sigma\left(\theta, \phi, \chi\right) \right)^2}}_{\text{aberration}} \underbrace{\frac{1}{\gamma\left( 1 - v \cos\sigma\left(\theta, \phi, \chi\right) \right)}}_{\text{time dilation}} \nonumber\\
&\qquad {} \times \delta\left( \frac{\omega'}{\underbrace{\gamma \left( 1 - v\cos\sigma\left(\theta, \phi, \chi\right) \right)}_{\text{relativistic shift}} \underbrace{\left(1+z\right)}_{\text{cosmological shift}}} - \omega \right) \nonumber\\
&\qquad {} \times \delta\left( \tau - r\left( \frac{1}{v} - \cos\sigma\left(\theta, \phi, \chi\right) \right) \left(1+z\right) \right)
\end{align}
Here $A_\text{det}$ is the effective area of the detector, and $A_\text{ph}$ is an area of the sphere over which photons emitted in the burst are spread (see appendix \ref{sec:cosmology} for derivation). We have taken account for four relativistic effects, which affect intensities and frequencies of the radiators: relativistic aberration; time dilation of the radiators relative to the observer; relativistic blue/redshift; and cosmological redshift.

@@

In the first test we compute the total energy radiated from the burst,
and to ensure that it doesn't exceed the mass of the star from which
the burst originated.

@@

For the modeling we first need to define the range of redshifts and
off-axis angles. The range should include all the observable jet
positions, but one should keep the range as small as possible in order
to avoid production of too many points corresponding to an invisible
bursts. For simplicity we consider the rectangular region.

@@

First of all, we need to determine, which parameters to fit. At most,
the observables of a particular burst depend on 10 parameters: 8 of
the burst $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k,
\alpha\right)$, and 2 of the observers position related to the burst
$\left(z, \chi\right)$. However, some of these parameters only change
observables trivially, and, also, there are transformations of
variables, which do not change observables at all (see equation
\ref{eq:p}):
\begin{itemize}
	\item{The number of photons $p$ depends linearly on $\eta_0$,
          so $\eta_0$ doesn't affect other observables like duration,
          or stretching factor. One can easily fit $\eta_0$ to match
          the observed photon count.}
	\item{If we make a simultaneous transformation of $r_0
          \rightarrow \lambda r_0$ and $\eta_0 \rightarrow
          \frac{1}{\lambda^3}\eta_0$, where $\lambda$ is a parameter
          of transformation, only duration of the burst will
          change. Other observables, like the total number of photons,
          or stretching factor, will not be affected.}
	\item{Finally, if we make a following transformation:
          $\omega_0 \rightarrow \lambda \omega_0$, $\theta_0
          \rightarrow \lambda^k \theta_0$, $\eta_0 \rightarrow
          \lambda^\alpha \eta_0$, where $\lambda$ is another
          transformation parameter, no observables will change
          whatsoever.}
\end{itemize}
Given these transformations, one can reduce the number of parameters
for fit to 7: $\left(\gamma, n, \theta_0, k, \alpha, z, \chi\right)$.

@@

First, the time stretching of GRB light curves between different VHE
(in particular $\unit[100]{MeV} < E < \unit[1]{GeV}$ and $E >
\unit[1]{GeV}$ ) bands is discovered with the statistical significance
of $3.3\sigma$.  Depending on the burst the stretching factor may be
higher of lower than 1, that is the higher or the lower-energy light
curve is stretched.

>>>>

We start with the location, time and duration of bursts from the
Fermi-LAT Gamma-Ray Burst Catalog \cite{Ackermann:2013zfa}. The
photons during the burst and the photons for 24 hours before the burst
are downloaded from the LAT Data
Server~\footnote{\url{http://fermi.gsfc.nasa.gov/cgi-bin/ssc/LAT/LATDataQuery.cgi}}.
The latter are used for the background estimation in both energy bands
with the technique introduced in \cite{Rubtsov:2011qq}. In order to
have enough statistics we require that at least 10 photons are
detected with the energy greater than $\unit[1]{GeV}$. This leaves us
with the four bursts for the time span of the Fermi-LAT Gamma-Ray
Burst Catalog. Namely, GRB~080916C \cite{Tajima:2009az}, GRB~090510
\cite{Ackermann:2010us}, GRB~090902B \cite{Abdo:2009pg} and
GRB~090926A \cite{Bregeon:2011bu}. For these bursts we stretch the
high energy light curve by the arbitrary factor and compare with the
low energy light curve using the Kolmogorov-Smirnov (KS) test. Within
the $95\%$ confidence level the stretching factor is compatible to $1$
(no stretching) for the GRB 080916C and GRB 090510. For the other two
bursts the stretching factor of 1 is, however, excluded.  For GRB
090902B it should be smaller than 1, so the low energy light curve is
stretched with respect to the high energy one, see
fig. \ref{fig:grb090902B}.  Even more significant deviation from the
stretching factor of 1 in the opposite direction is observed for the
GRB 090926A. The high energy light curve of GRB 090926A is stretched
with respect to the low energy one by a factor of at least 1.99 (see
fig. \ref{fig:grb090926A}).  The stretching of the GRB 090926A high
energy light is found with $3.3\sigma$ significance.  All the observed
stretching factors are summarized in Table
\ref{tab:observationResults}, and the detailed procedure for their
calculation is described in Section \ref{sec:observations}.

@@

First, we download the Pass 7REP (V15) ``SOURCE'' class photons from
the Fermi-LAT data
server~\cite{2009ApJ...697.1071A,Ackermann:2012kna}. In order to
define the regions of interest we use the catalog
\cite{Ackermann:2013zfa}, which contains all the bright gamma-ray
bursts seen by the Fermi LAT since 2008 August 4 to 2011 August 1.
Specifically, we take 4 pieces of information from the tables 2 and 4
of Ref. \cite{Ackermann:2013zfa}:
\begin{itemize}
	\item{
		GBM trigger time.
		It is used as a reference point for the GRB time.
	}
	\item{
		Location of the GRB.
		The burst locations are used to filter out photons
                coming from the other sources.
	}
	\item{
		Location error.
		Location errors are used to improve the accuracy of filtering, specifically to avoid losing statistics by filtering too much.
	}
	\item{
		$T_{05}$ -- $T_{95}$ interval of the LAT-detected
          emission. The beginning of the interval is used as a fixed
          point of stretching. The Fermi LAT photons are downloaded
          for the period extended in time by $50\%$ towards both past
          and future compared to this interval.
	}
\end{itemize}
The data are summarized in table \ref{tab:bursts}. For the background
estimation, we also download observational data for 24 hours before
the burst.  The analysis is performed with the Fermi Science Tools
v9r33p0
package~\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/scitools/overview.html}}
following the
guidelines~\footnote{\url{http://fermi.gsfc.nasa.gov/ssc/data/analysis/scitools/data_preparation.html}}.
We require Earth zenith angle to be below $100^\circ$ to remove the
Earth limb emission. After the basic filtering by \texttt{gtselect}
and \texttt{gtmktime} tools is done we perform more elaborate
selection using the point spread function of Fermi LAT, calculated
with \texttt{gtpsf} tool. We use 300 energy bins from \unit[100]{MeV}
to \unit[300]{GeV}, and 300 angular bins with maximal angle being
$30\degree$.  The \texttt{gtltcube} tool is used to compute required
livetime maps. Note, that we calculate PSFs separately for both
conversion types: front and back. We define $\mathrm{PSF}_{95}$ as an
angular distance from the source containing 95\% of the emitted
photons. We require that the photon is closer than $\mathrm{PSF}_{95}
+ \texttt{location.err}$ to the location of the GRB (see Table
\ref{tab:bursts}). The details of the data request and
analysis are given in appendix \ref{sec:fermiCode}.

@@

Although the details of the exact mechanism of the GRB are not known
there are arguments in support of our model.

@@

\subsection{Light curve}
We now know enough to compute the observable quantity -- the number of
observed photons $p$. For that we need to integrate the radiation
intensity $\eta$ over the four groups of variables. Two groups are
burst-related: radiator position $\left(r, \theta, \phi\right)$ and
frequency at emission $\omega'$. The other two variables are related
to observer: observation time $\tau$ and frequency at observation
$\omega$. We relate the first two and the second two with the delta
functions:
\begin{align}
p\left( z, \chi; \tau_1, \tau_2; \omega_1, \omega_2 \right) &= \frac{A_\text{det}}{A_\text{ph}\left(z\right)} \int_0^\infty \dif r \int_0^{\pi} r \dif \theta \int_0^{2\pi} r \sin{\theta} \dif \phi \int_0^\infty \dif \omega' \int_{\tau_1}^{\tau_2} \dif \tau \int_{\omega_1}^{\omega_2} \dif \omega \, \eta\left( r, \theta, \omega' \right) \nonumber\\
&\qquad {} \times \underbrace{\frac{1}{\gamma^2\left( 1 - v \cos\sigma\left(\theta, \phi, \chi\right) \right)^2}}_{\text{aberration}} \underbrace{\frac{1}{\gamma\left( 1 - v \cos\sigma\left(\theta, \phi, \chi\right) \right)}}_{\text{time dilation}} \nonumber\\
&\qquad {} \times \delta\left( \frac{\omega'}{\underbrace{\gamma \left( 1 - v\cos\sigma\left(\theta, \phi, \chi\right) \right)}_{\text{relativistic shift}} \underbrace{\left(1+z\right)}_{\text{cosmological shift}}} - \omega \right) \nonumber\\
&\qquad {} \times \delta\left( \tau - r\left( \frac{1}{v} - \cos\sigma\left(\theta, \phi, \chi\right) \right) \left(1+z\right) \right)
\end{align}
Here $A_\text{det}$ is the effective area of the detector, and $A_\text{ph}$ is an area of the sphere over which photons emitted in the burst are spread (see appendix \ref{sec:cosmology} for derivation). We have taken account for four relativistic effects, which affect intensities and frequencies of the radiators: relativistic aberration; time dilation of the radiators relative to the observer; relativistic blue/redshift; and cosmological redshift.

@@

In the first test we compute the total energy radiated from the burst,
and ensure that it doesn't exceed the mass of the star from which
the burst originated.

@@

For the modeling we first need to define the range of redshifts and
off-axis angles. The range should include all the observable jet
positions, but one should keep the range as small as possible in order
to avoid production of too many points corresponding to invisible
bursts. For simplicity we consider the rectangular region.

@@

First of all, we need to determine, which parameters to fit. At most,
the observables of a particular burst depend on 10 parameters: 8 of
the burst $\left(\gamma, \eta_0, r_0, n, \omega_0, \theta_0, k,
\alpha\right)$, and 2 of the observer's position related to the burst
$\left(z, \chi\right)$. However, some of these parameters only change
observables trivially, and, also, there are transformations of
variables, which do not change observables at all (see equation
\ref{eq:p}):
\begin{itemize}
	\item{The number of photons $p$ depends linearly on $\eta_0$,
          so $\eta_0$ doesn't affect other observables like duration,
          or stretching factor. One can easily fit $\eta_0$ to match
          the observed photon count.}
	\item{If we make a simultaneous transformation of $r_0
          \rightarrow \lambda r_0$ and $\eta_0 \rightarrow
          \frac{1}{\lambda^3}\eta_0$, where $\lambda$ is a parameter
          of transformation, only duration of the burst will
          change. Other observables, like the total number of photons,
          or stretching factor, will not be affected.}
	\item{Finally, if we make a following transformation:
          $\omega_0 \rightarrow \lambda \omega_0$, $\theta_0
          \rightarrow \lambda^k \theta_0$, $\eta_0 \rightarrow
          \lambda^\alpha \eta_0$, where $\lambda$ is another
          transformation parameter, no observables will change
          whatsoever.}
\end{itemize}
Given these transformations, one can reduce the number of parameters
for fit to 7: $\left(\gamma, n, \theta_0, k, \alpha, z, \chi\right)$.

@@

First, the time stretching of GRB light curves between different VHE
(in particular $\unit[100]{MeV} < E < \unit[1]{GeV}$ and $E >
\unit[1]{GeV}$) bands is discovered with the statistical significance
of $3.3\sigma$.  Depending on the burst the stretching factor may be
higher or lower than 1, that is the higher or the lower-energy light
curve is stretched.
